Inattentive responding can induce spurious associations between task behavior and symptom measures

### 2021-04-11 manuscripts/main.tex
ready for submission

% make abstract
\abstract{
A common research design in the field of computational psychiatry involves leveraging the power of online participant recruitment to assess correlations between behaviour in cognitive tasks and the self-reported severity of psychiatric symptoms in large, diverse samples. Although large online samples have many advantages for psychiatric research, some potential pitfalls of this research design are not widely understood. Here we detail circumstances in which entirely spurious correlations may arise between task behaviour and symptom severity as a result of inadequate screening of careless or low-effort responding on psychiatric symptom surveys. Specifically, since many psychiatric symptom surveys have asymmetric ground-truth score distributions in the general population, participants who respond carelessly on these surveys will show apparently elevated symptom levels. If these participants are similarly careless in their task performance, and are not excluded from analysis, this may result in a spurious association between greater symptom scores and worse behavioural task performance. Here, we demonstrate exactly this pattern of results in $N = 386$ participants recruited online to complete a self-report symptom battery and a short reversal-learning choice task. We show that many behaviour-symptom correlations are entirely abolished when participants flagged for careless responding on surveys are excluded from analysis. We also show that exclusion based on task performance alone is not sufficient to prevent these spurious correlations. Of note, we demonstrate that false-positive rates for these spurious correlations \emph{increase} with sample size, contrary to common assumptions. We offer guidance on how researchers using this general experimental design can guard against this issue in future research; in particular, we recommend the adoption of screening methods for self-report measures that are currently uncommon in this field.
}

@@

In recent years, online labour markets (e.g. Amazon Mechanical Turk, Prolific, Crowdflower) have become increasingly popular as a source of research participants in the behavioral sciences \cite{stewart2017crowdsourcing}, in no small part due to the ease with which these services allow for recruitment of large, diverse samples. The advantages of online data collection have also begun to be recognized in psychiatric research \cite{chandler2016conducting}, where this method offers several distinct advantages over traditional approaches to participant recruitment. The ability to assess psychiatric symptom severity in large general-population samples makes possible large-scale transdiagnostic analysis \cite{gillan2016taking, rutledge2019machine}, and facilitates recruitment from difficult-to-reach participant populations \cite{strickland2019use}. Online labour markets also facilitate re-recruitment, making them an attractive option for validating the psychometric properties of assessment tools \cite{enkavi2019large} or studying clinical processes longitudinally \cite{kothe2019retention}.

@@

When participants complete both symptom surveys and behavioral tasks---a common study design in computational psychiatry---this artefact has the potential to induce spurious correlations between symptom self-reports and task behavior. That is, while C/IE behavior is traditionally thought of as a source of noise that can result in type II (false negative) errors, here we suggest that in large-scale online psychiatric studies it can instead result in type I (false positive) errors. Concretely, if the same participants who engage in C/IE responding on surveys (and who therefore inaccurately report high levels of psychiatric symptoms) also respond with insufficient effort on behavioral tasks, this can cause experimenters to observe an entirely spurious correlation between greater symptom severity and worse task performance (see Figure \ref{fig:simulation}). A similar effect has been well documented in personality psychology, where the presence of C/IE responding can induce correlations between questionnaires, and bias estimated factors in factor analysis \cite{huang2012detecting, robinson2014inaccurate, huang2015insufficient, chandler2020participant, arias2020little}.

\begin{figure}[t]
\includegraphics[width=16cm]{../figures/main_01.png}
\centering
\captionsetup{width=0.88\textwidth}
\caption{Simulated example of how spurious behavior-symptom correlations can arise when symptom endorsement is rare. \textit{Left:} When symptoms are moderately common in the general population, C/IE respondents (blue) are indistinguishable from attentive participants (red) in self-report measures (x-axis, marginal distribution shown on top). Despite the lower task performance of C/IE respondents (y-axis), no correlation arises between symptom scores and task performance (dots are participants drawn from the shown distributions, with \~15\% C/IE participants; dashed line shows the (lack of) correlation.) \textit{Right:} When symptoms are rare in the general population, careless respondents appear symptomatic in self-report measures. As a result, self-report symptom scores show a significant correlation with task performance.}
\label{fig:simulation}
\end{figure}

@@

Taken together, these results suggest that measures of C/IE responding in task and self-report data do not identify the same set of participants. This means that solely excluding participants on the basis of poor behavioural performance---the most common approach in online studies---is unlikely to identify participants who engage in C/IE responding on self-report surveys.

@@

In this manuscript, we highlighted a particular set of circumstances, common in computational psychiatry research done on large online samples, in which spurious correlations may arise between task behavior and self-reported symptomology. When the ground-truth prevalence of a symptom is low in the general population, participants who respond carelessly on measures assessing this symptom may erroneously appear as symptomatic. A less careful pattern of responding on tasks used to measure cognitive constructs can then masquerade as a correlation between alteration in these constructs and symptom dimensions. We found empirical evidence for this pernicious pattern in participants recruited from two popular online labour platforms. False-positive rates for these spurious correlations \emph{increased} with sample size, because the correlations are due to measurement bias, not measurement noise. Importantly, we found that screening on task behavior alone was insufficient to identify participants engaging in C/IE responding and avoid the false-positive correlations. Unfortunately, a literature review identified this type of screening as the most common practice in online computational psychiatry studies. We recommend instead to screen and exclude participants based on responding on surveys, a practice that abolished many spurious behaviour-symptom correlations in our data.

@@

We fit the softmax regression model using Stan (details above).Participant parameters were fit individually (i.e. not hierarchically)  so as to prevent bias during parameter estimation from partial-pooling between attentive and C/IE participants. Parameters were sampled with Gaussian priors with $\mu = 0$ and $\sigma = 5$. 
\section{Acknowledgements}

The authors are grateful to Agnes Norbury, Alexandra Pike, and Oliver Robinson for helpful discussion. This work was funded in part by NIMH 5R01MH119511-02 and ., SZ was supported by an NSF Graduate Research Fellowship. DB was supported by an Early Career Fellowship from the Australian National Health and Medical Research Council (\#1165010).

>>>>

% make abstract
\abstract{
A common research design in the field of computational psychiatry involves leveraging the power of online participant recruitment to assess correlations between behavior in cognitive tasks and the self-reported severity of psychiatric symptoms in large, diverse samples. Although large online samples have many advantages for psychiatric research, some potential pitfalls of this research design are not widely understood. Here we detail circumstances in which entirely spurious correlations may arise between task behavior and symptom severity as a result of inadequate screening of careless or low-effort responding on psychiatric symptom surveys. Specifically, since many psychiatric symptom surveys have asymmetric ground-truth score distributions in the general population, participants who respond carelessly on these surveys will show apparently elevated symptom levels. If these participants are similarly careless in their task performance, and are not excluded from analysis, this may result in a spurious association between greater symptom scores and worse behavioral task performance. Here, we demonstrate exactly this pattern of results in $N = 386$ participants recruited online to complete a self-report symptom battery and a short reversal-learning choice task. We show that many behavior-symptom correlations are entirely abolished when participants flagged for careless responding on surveys are excluded from analysis. We also show that exclusion based on task performance alone is not sufficient to prevent these spurious correlations. Of note, we demonstrate that false-positive rates for these spurious correlations \emph{increase} with sample size, contrary to common assumptions. We offer guidance on how researchers using this general experimental design can guard against this issue in future research; in particular, we recommend the adoption of screening methods for self-report measures that are currently uncommon in this field.
}

@@

In recent years, online labour markets (e.g. Amazon Mechanical Turk, Prolific, Cloudresearch) have become increasingly popular as a source of research participants in the behavioral sciences \cite{stewart2017crowdsourcing}, in no small part due to the ease with which these services allow for recruitment of large, diverse samples. The advantages of online data collection have also begun to be recognized in psychiatric research \cite{chandler2016conducting}, where this method offers several distinct advantages over traditional approaches to participant recruitment. The ability to assess psychiatric symptom severity in large general-population samples makes possible large-scale transdiagnostic analysis \cite{gillan2016taking, rutledge2019machine}, and facilitates recruitment from difficult-to-reach participant populations \cite{strickland2019use}. Online labour markets also facilitate re-recruitment, making them an attractive option for validating the psychometric properties of assessment tools \cite{enkavi2019large} or studying clinical processes longitudinally \cite{kothe2019retention}.

@@

When participants complete both symptom surveys and behavioral tasks---a common study design in computational psychiatry---this artifact has the potential to induce spurious correlations between symptom self-reports and task behavior. That is, while C/IE behavior is traditionally thought of as a source of noise that can result in type II (false negative) errors, here we suggest that in large-scale online psychiatric studies it can instead result in type I (false positive) errors. Concretely, if the same participants who engage in C/IE responding on surveys (and who therefore inaccurately report high levels of psychiatric symptoms) also respond with insufficient effort on behavioral tasks, this can cause experimenters to observe an entirely spurious correlation between greater symptom severity and worse task performance (see Figure \ref{fig:simulation}). A similar effect has been well documented in personality psychology, where the presence of C/IE responding can induce correlations between questionnaires, and bias estimated factors in factor analysis \cite{huang2012detecting, robinson2014inaccurate, huang2015insufficient, chandler2020participant, arias2020little}.

\begin{figure}[t]
\includegraphics[width=16cm]{../figures/main_01.png}
\centering
\captionsetup{width=0.88\textwidth}
\caption{Simulated example of how spurious behavior-symptom correlations can arise when symptom endorsement is rare. \textit{Left:} When symptoms are moderately common in the general population, C/IE respondents (blue) are indistinguishable from attentive participants (red) in self-report measures (x-axis, marginal distribution shown on top). Despite the lower task performance of C/IE respondents (y-axis), no correlation arises between symptom scores and task performance (dots are participants drawn from the shown distributions, with 15\% C/IE participants; dashed line shows the (lack of) correlation.) \textit{Right:} When symptoms are rare in the general population, careless respondents appear symptomatic in self-report measures. As a result, self-report symptom scores show a significant correlation with task performance.}
\label{fig:simulation}
\end{figure}

@@

Taken together, these results suggest that measures of C/IE responding in task and self-report data do not identify the same set of participants. This means that solely excluding participants on the basis of poor behavioral performance---the most common approach in online studies---is unlikely to identify participants who engage in C/IE responding on self-report surveys.

@@

In this manuscript, we highlighted a particular set of circumstances, common in computational psychiatry research done on large online samples, in which spurious correlations may arise between task behavior and self-reported symptomology. When the ground-truth prevalence of a symptom is low in the general population, participants who respond carelessly on measures assessing this symptom may erroneously appear as symptomatic. A less careful pattern of responding on tasks used to measure cognitive constructs can then masquerade as a correlation between alteration in these constructs and symptom dimensions. We found empirical evidence for this pernicious pattern in participants recruited from two popular online labour platforms. False-positive rates for these spurious correlations \emph{increased} with sample size, because the correlations are due to measurement bias, not measurement noise. Importantly, we found that screening on task behavior alone was insufficient to identify participants engaging in C/IE responding and avoid the false-positive correlations. Unfortunately, a literature review identified this type of screening as the most common practice in online computational psychiatry studies. We recommend instead to screen and exclude participants based on responding on surveys, a practice that abolished many spurious behavior-symptom correlations in our data.

@@

We fit the softmax regression model using Stan (details above).Participant parameters were fit individually (i.e. not hierarchically)  so as to prevent bias during parameter estimation from partial-pooling between attentive and C/IE participants. Parameters were sampled with Gaussian priors with $\mu = 0$ and $\sigma = 5$.

\section{Acknowledgements}

The authors are grateful to Agnes Norbury, Alexandra Pike, and Oliver Robinson for helpful discussion. The research reported in this manuscript was supported in part by the National Institute of Mental Health (NIMH) under award number 5R01MH119511-02, and by the National Center for Advancing Translational Sciences (NCATS), a component of the National Institute of Health (NIH), under award number UL1TR003017. The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health. SZ was supported by an NSF Graduate Research Fellowship. DB was supported by an Early Career Fellowship from the Australian National Health and Medical Research Council (\#1165010).
